Investigation of thermal reactionomes reveals stress tolerance vs. resistance in

### 2014-08-22 doc/ApTranscriptome_manuscript.Rmd
edits to Intro

Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009]. Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, varies considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011]. Outside of candidate genes [@krebs1998], little is known about the genetic basis of this variation or how it has evolved. To address this, we need information on the extent to which geographic differentiation and selection has acted upon the diversity of genes, and their plasticity of expression, involved in thermal tolerance. This information is a critical for understanding the adaptive potential of species to future climates [@hoffmann2008] and the link between short term acclimation and adaptation to increased temperatures [@somero2010].

@@

Here, we present a novel method for RNAseq analysis: the *thermal reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across the temperatures it experiences. We use this method to identify genes that are thermally-responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996], and then compare the thermal reactionomes of these species to gain insight into how spatially divergent selection has led to changes in thermally-responsive gene expression.

Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006; @wardle2010]. Temperature is a potentially strong selective force for ground-nesting ant populations, which must tolerate seasonally freezing winters and hot summers. On shorter time scales, individual workers can experience extreme thermal environments when they leave the thermally buffered ant nest to forage for food [@diamond2012]. Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012], and species as well as populations within species differ in the critical maximum and minimum temperatures [@warren2013].

As the more southern *A. carolinensis* occurs in climates that are hotter with longer growing seasons than that of the more northerly *A. picea*, we expected patterns of gene expression in response to thermal extremes to differ between these species in a number of ways. First, the number of genes showing plasticity of expression to thermal extremes may differ between the species. Finding that more genes are up-regulated to high temperatures in the colder-climate species *A. picea* would indicate that this species relies on activation of the cellular stress response [@kultz2005] and other protective genes to a greater extent than *A. carolinensis*, while the opposite would be true for cold temperatures. Conversely, finding that more genes are down-regulated in response to thermal extremes would indicate that a species relies to a greater extent upon proteome stability [@ghosh2010] and behavioral mechanisms [@sunday2014] to tolerate thermal stress.

Second, responsive genes may change in their sensitivity to thermal extremes. Finding that heat-responsive genes are up-regulated at higher temperatures in *A. carolinensis* than in *A. picea* would be consistent with local adaptation for greater heat tolerance. That is, more extreme temperatures are required for activation of the heat shock response [@somero2011]. Conversely, finding cold-responsive genes are up-regulated at lower temperatures in *A. picea* than *A. carolinensis* would indicate local adaptation for cold tolerance. Similar patterns may be observed for genes that are down-regulated in response to thermal extremes. Finally, changes in the expression patterns of responsive genes between species can provide insight on the evolution of thermal tolerance. Specifically, if adaptation to thermal extremes works by genetic assimilation [@pigliucci2006], genes up-regulated in response to cold temperatures in *A. carolinensis* may have increased basal expression in the colder-climate species *A. picea*, and vice versa.

@@

>>>>

Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009]. Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, varies considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011]. Outside of candidate genes [@krebs1998], little is known about the genetic basis of this variation or how it has evolved. To address this, we need information on the extent to which geographic differentiation and selection has acted upon the diversity of genes, and their plasticity of expression, involved in thermal tolerance. This information is a critical for understanding the adaptive potential of species to future climates [@hoffmann2008] and the link between short-term physiological acclimation and longer-term adaptation to increased temperatures [@somero2010].

@@

Here, we present a novel method for RNAseq analysis: the *thermal reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across the temperatures it experiences. We use this method to identify genes that are thermally-responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996]. Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006; @wardle2010].

We compared the thermal reactionomes of these species to gain insight on the relationship between short-term physiological acclimation and longer-term adaptive changes to temperature. First, through the process of genetic assimilation [@pigliucci2006], adaptive responses are hypothesized to involve the same molecular pathways as short-term acclimatory responses. Finding that the same genes up-regulated in response to cold temperatures in *A. carolinensis* have increased basal expression in the colder-climate species *A. picea*, and vice versa, would be consistent with this pattern.

Second, spatially-divergent adaptive responses to temperature may involve changes

the molecular toolkit involved in short-term acclimation to temperature is the same as

finding that more genes are up-regulated to high temperatures in the colder-climate species *A. picea* would indicate that this species relies on activation of the cellular stress response [@kultz2005] and other protective genes to a greater extent than *A. carolinensis*, which might rely to a greater extent on proteome stability [@ghosh2010] and behavioral avoidance mechanisms [@sunday2014], while the opposite would be true for cold temperatures.

Second, responsive genes may change in their sensitivity to thermal extremes. Finding that heat-responsive genes are up-regulated at higher temperatures in *A. carolinensis* than in *A. picea* would be consistent with local adaptation for greater heat tolerance. That is, more extreme temperatures are required for activation of the heat shock response [@somero2011]. Conversely, finding cold-responsive genes are up-regulated at lower temperatures in *A. picea* than *A. carolinensis* would indicate local adaptation for cold tolerance. Similar patterns may be observed for genes that are down-regulated in response to thermal extremes. Finally, changes in the expression patterns of responsive genes between species can provide insight on the evolution of thermal tolerance. Specifically, if adaptation to thermal extremes works by genetic assimilation [@pigliucci2006], genes

@@

...ants...Temperature is a potentially strong selective force for ground-nesting ant populations, which must tolerate seasonally freezing winters and hot summers. On shorter time scales, individual workers can experience extreme thermal environments when they leave the thermally buffered ant nest to forage for food [@diamond2012]. Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012], and species as well as populations within species differ in the critical maximum and minimum temperatures [@warren2013].
